```{r}
library(tidyverse)
library(postmastr)
library(usdata)
library(rvest)
```

# Load the Source Data

```{r, message=FALSE}
wbb_rosters23 <- read_csv("https://raw.githubusercontent.com/dwillis/wbb-rosters/main/rosters_2022-23.csv")
```

# Teams That Have Roster Data

```{r, message=FALSE}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/teams.csv") %>% select(-url, -twitter, -stats_name, -private)

roster_teams <- wbb_rosters23 %>% distinct(ncaa_id) %>% mutate(has_roster = 1)

teams <- teams %>% 
  left_join(roster_teams) %>% mutate(roster = has_roster)

teams_without_rosters <- teams %>% 
  filter(is.na(roster)) %>% 
  filter(!is.na(division))

wbb_rosters23 <- wbb_rosters23 %>% 
  left_join(teams, by=c("ncaa_id", "team")) 
```

# Height Cleaning

Describe the source data and what you needed to do to make it usable, and then what you did to make it possible to standardize heights.

```{r}
wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(height_clean = str_replace(height, "'", '-')) %>% 
  mutate(height_clean = str_replace(height_clean, "’", '-')) %>% 
  mutate(height_clean = str_replace(height_clean, "''", '')) %>% 
  mutate(height_clean = str_replace(height_clean, '"', '')) %>%
  separate(height_clean, c('height_ft', 'height_in'), sep="-", extra="merge") %>% 
  mutate(height_ft = as.numeric(height_ft), height_in = as.numeric(height_in)) %>%
  mutate(total_inches = (height_ft*12)+height_in)
```

# Positions Cleaning

Describe the source data and what you needed to do to make it usable, and then what you did to make it possible to standardize the positions.

```{r}
positions_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/wbb_positions.csv") %>%
  select(-count)

# replace Ball State's position column with NAs
wbb_rosters23 <- wbb_rosters23 %>% mutate(position = ifelse(ncaa_id == 47, NA, position))

# need to fix rows where the height is also in the position column
# in many of those cases we used information from biography pages and/or previous teams or news coverage to assign the position.

position_fixes <- wbb_rosters23 %>% left_join(positions_cleaned, by = "position") %>% filter(position == height)

wbb_rosters23 <- wbb_rosters23 %>% left_join(positions_cleaned, by = "position")

wbb_rosters23 %>% distinct(position_clean)

## Primary/Secondary positions based on order of position_clean

wbb_rosters23 <- wbb_rosters23 %>% mutate(position_full = position_clean)
wbb_rosters23 <- wbb_rosters23 %>% mutate(position_clean = na_if(position_clean, "N/A"))
wbb_rosters23 <- wbb_rosters23 %>% separate(position_clean, c('primary_position', 'secondary_position'), sep='/', extra='merge')
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position)) 
wbb_rosters23 <- wbb_rosters23 %>% mutate(secondary_position = case_when((secondary_position == "G") ~ "Guard", (secondary_position == "C") ~ "Center", (secondary_position == "F") ~ "Forward", TRUE ~ secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% mutate(primary_position = case_when((primary_position == "G") ~ "Guard", (primary_position == "C") ~ "Center", (primary_position == "F") ~ "Forward", TRUE ~ primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$primary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("'", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("5", wbb_rosters23$secondary_position))
wbb_rosters23 <- wbb_rosters23 %>% filter(!grepl("6", wbb_rosters23$secondary_position))

```

# Years Cleaning

Describe the source data and what you needed to do to make it usable.

12/6 - Need to check NA year_clean and regenerate yearscleaned.csv with new entries from missing_years. Need to update ~33 rows:

Liberty
https://golobos.com/sports/wbball/roster/season/2022-23/player/amaya-brown/ (RS-Sr.)
https://www.valpoathletics.com/wbasketball/roster/2022-23/12601/ilysse-pitts/ (5th)
https://csurams.com/sports/womens-basketball/roster/petra-farkas/8848 & https://ysusports.com/sports/womens-basketball/roster/lindsey-mack/6141 (Fifth Year)
https://lsusports.net/sports/wbball/roster/player/alexis-morris/ (5th-year Senior)


```{r}
years_cleaned <- read_csv("https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/yearscleaned.csv")

missing_years <- left_join(wbb_rosters23, years_cleaned) %>% rename(year_clean = `year-clean`) %>% filter(is.na(year_clean))

wbb_rosters23 <- left_join(wbb_rosters23, years_cleaned) %>% rename(year_clean = `year-clean`) %>% select(-count)

```

# Previous School Cleaning

Some records list the player's high school in the previous school column. We filter those out and turn it into a list, then update `wbb_rosters23` to replace hs_clean with the previous school for those rows.

```{r}
hs_wbb <- wbb_rosters23 %>%
  filter(is.na(high_school))%>%
  filter(!is.na(previous_school))%>%
  filter(str_detect(previous_school, "HS"))%>%
  mutate(hs_clean = "")%>%
  mutate(hs_clean = previous_school)%>%
  mutate(type = "has hs")
extras_wbb <- wbb_rosters23 %>%
  filter(!is.na(high_school))%>%
  filter(is.na(previous_school))%>%
  mutate(hs_clean = "")%>%
  mutate(type = "no hs")
all_wbb <- bind_rows(hs_wbb, extras_wbb)

previous_school_is_hs=as.list(hs_wbb$previous_school)

# update based on list membership

wbb_rosters23 <- wbb_rosters23 %>%
  mutate(hs_clean = case_when(
    is.na(high_school) & previous_school %in% previous_school_is_hs ~ previous_school,
    TRUE ~ high_school)
  )

# where previous school == hs_clean, replace previous_school with NA

wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(previous_school = ifelse(previous_school == hs_clean, NA, previous_school))
```

# Hometown/Homestate Cleaning

The original data for the homestate column comes from the hometown. We clean up these strings to standardize them before we split them up and use the package 'postmastr' to look for state abbreviations or spellings. If the package finds a state, then great. If not, we check for any weird spellings/abbreviations of state and replace it with a standardized abbreviation. In the end, if the player is from the US, we should have a postal state abbreviation. If not, then we should have <NA>

```{r}
wbb_rosters23 <- wbb_rosters23 %>% 
  mutate(hometown_cleaned = str_to_upper(str_replace(hometown, "\\.",""))) %>% 
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"/.*","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"\\.$","")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,",",", ")) %>%
  mutate(hometown_cleaned = str_replace(hometown_cleaned,"-"," ")) %>% 
  mutate(hometown_cleaned = str_squish(hometown_cleaned))

states_non_standard <- pm_append(type = "state", input = c("SD.", "MASS", "CALIF", "MICH", "NEB", "IND", "MINN", "ORE", "OHIO", "FLA", "MISS", "TENN", "ARIZ", "KAN", "ALA", "OKLA", "WIS", "ILL", " WASH", "ARK", "COLO", "NEV", "CONN", "WISC", "WVA", "DEL", "WYO", "CALI", "LOUIS", "VIRG", "MONT", "PENN", "TEX", "KANS", "NEBR", "IDA", "COL"), output = c("SD", "MA", "CA", "MI", "NE", "IN", "MN", "OR", "OH", "FL", "MS", "TN", "AZ", "KS", "AL", "OK", "WI", "IL", "WA", "AR", "CO", "NV", "CT", "WI", "WV", "DE", "WY", "CA", "LA", "VA", "MT", "PA", "TX", "KS", "NE", "ID", "CO"), locale = "us")

dict <- pm_dictionary(type='state', case = c("title", "upper", "lower"), append = states_non_standard)

wbb_rosters23 <- wbb_rosters23 %>% pm_identify(var="hometown_cleaned")

parsed <- wbb_rosters23 %>% 
  pm_identify(var="hometown_cleaned") %>% 
  pm_prep(var="hometown_cleaned", type="street") %>% 
  pm_state_parse(dict)

wbb_rosters23 <- left_join(wbb_rosters23, parsed, by="pm.uid")

wbb_rosters23 <- wbb_rosters23 %>% separate(hometown, c('hometown', 'homestate'), sep=',', extra='merge')

wbb_rosters23 <- wbb_rosters23 %>% mutate(homestate=str_trim(homestate), homestate=str_replace(homestate, '\\.', '')) %>%
  mutate(homestate = case_when(is.na(abbr2state(homestate)) ~ homestate, TRUE ~ (abbr2state(homestate))))

# fix Hawai'i & some others

wbb_rosters23 <- wbb_rosters23 %>%
  mutate(pm.state = case_when(
    homestate %in% c("Hawai‘i", "O‘ahu") ~ "HI",
    homestate == 'Ken' ~ 'KY',
    homestate == 'Illnois' ~ 'IL',
    homestate == 'Republic of Georgia' ~ NA,
    homestate == 'N C.' ~ 'NC',
    homestate == 'Oreg' ~ 'OR',
    homestate == 'US.V.I.' ~ 'USVI',
    homestate %in% c('Calfiornia', 'Califonia') ~ 'California',
    TRUE ~ pm.state
  ))

```

# Foreign Nation Cleaning

To extract the player's country, we are using a list of FIBA nations as a starting point. We're splitting the hometown column into multiple columns and seeing if we can detect any of the FIBA nations in the last column, where the country should be. In addition to this, we are also using some conditional statements in our code to handle special cases where the country is either not in FIBA or there are spelling errors/inconsistencies in the way countries are listed. Through this process, if a country is detected in the original hometown string, we fill a new column for the players country.

The values left are vast majority cities, either American or not, that are not listed with a state or country. For the sake of being conservative in our assumptions about where a city is located, we leave these rows' countries as <NA>.

```{r}
fibaurl <- "https://www.fiba.basketball/rankingwomen"
nations <- fibaurl %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="fiba_ranking_table_wrapper"]/table') %>%
  html_table()
nations_df <- nations[[1]]
wbb_rosters23 <- mutate(wbb_rosters23, country = case_when(!is.na(pm.state) ~ "USA"))

wbb_rosters23 <- wbb_rosters23 %>% mutate(temp = sapply(strsplit(wbb_rosters23$hometown_cleaned, ", ", fixed=TRUE), tail, 1))

wbb_rosters23 <- mutate(wbb_rosters23, country = case_when(
  temp %in% c('BC', 'QUEBEC', 'BRITISH COLUMBIA', 'ALBERTA', 'ONTARIO', 'NOVA SCOTIA', 'ONT', 'QUÉBEC', 'MANITOBA', 'QUE', 'ON', 'QC', 'ALBERTA (AB)', 'ONT CANADA', 'BC CANADA', 'BC,CANADA', 'ON CANADA') ~ "CANADA", 
  temp == "THE NETHERLANDS" ~ "NETHERLANDS", 
  temp == "SOUTH AUSTRALIA" ~ "AUSTRALIA",  
  temp == "TASMANIA" ~ "AUSTRALIA",
  temp == "AUSTRAILA" ~ "AUSTRALIA",
  temp == "WESTERN AUSTRALIA" ~ "AUSTRALIA",
  temp == "BOSNIA" ~ "BOSNIA AND HERZEGOVINA",
  temp == "MACEDONIA" ~ "MACEDONIA",
  temp == "TAIWAN" ~ "TAIWAN",
  temp == "PAPUA NEW GUINEA" ~ "PAPUA NEW GUINEA",
  temp == "WEST AFRICA" ~ "SENEGAL",
  temp == "UNITED KINGDOM" ~ "ENGLAND",
  temp == "UK" ~ "ENGLAND",
  temp == "HAWKS BAY" ~ "NEW ZEALAND",
  temp == "HAWAI'I" ~ "USA",
  temp == "SWEEDEN" ~ "SWEDEN",
  temp == "BELARUS" ~ "BELARUS",
  temp == "BANGLADESH" ~ "BANGLADESH",
  temp == "FRENCH GUIANA" ~ "FRENCH GUIANA",
  temp == "REPUBLIC OF GEORGIA" ~ "GEORGIA",
  temp == "DEMOCRATIC REPUBLIC OF THE CONGO" ~ "DEMOCRATIC REPUBLIC OF THE CONGO",
  "EUROPE" %in% temp ~ str_replace(temp, "EUROPE", "")
  TRUE ~ country))

nations_df$Country <- toupper(nations_df$Country)

added_nations <- data.frame(Worldrank=c(0, 0, 0, 0),
                  Country=c('ENGLAND', 'RUSSIA', 'SCOTLAND', 'NORTHERN IRELAND'),
                  Zonerank=c(0, 0, 0, 0),
                  IOC=c('', '', '', ''),
                  "Current points"=c(0, 0, 0, 0),
                  "+/- Rank *"=c(0, 0, 0, 0),
                  check.names = FALSE)
nations_df <- rbind(nations_df, added_nations)

wbb_rosters23 <- left_join(wbb_rosters23, nations_df, by=c('temp'='Country'))

wbb_rosters23 <- mutate(wbb_rosters23, country = case_when((!is.na(Worldrank) & is.na(country)) ~ temp, is.na(Worldrank) &  !is.na(country) ~ country, country == "USA" ~ "USA"))

wbb_rosters23 <- subset(wbb_rosters23, select = -c(Worldrank,Zonerank,IOC,`Current points`, `+/- Rank *`) )

wbb_rosters23 %>% filter(is.na(country)) %>% arrange(temp)
```
```{r}
wbb_rosters23 %>% group_by(hometown,homestate, country) %>% summarise(count=n()) %>% arrange(desc(count))
wbb_rosters23 <- subset(wbb_rosters23, select = -c(temp) )
```

# Rename Cleaned Columns, Remove Unneeded Columns and Write Out Final CSV File

```{r}
wbb_rosters23 <- subset(wbb_rosters23, select = -c(pm.id, pm.uid, pm.type, pm.address))

wbb_rosters23 <- wbb_rosters23 %>% 
  rename(height_clean = height, position_clean = position_full, previous_school_clean = previous_school, hometown_clean = hometown_cleaned, state_clean = pm.state, foreign_nation_clean = country)

write_csv(wbb_rosters23, "wbb_rosters_2022_23.csv")
```

