# Women's College Basketball Roster Data Cleaning

This notebook processes and standardizes women's college basketball roster data for a single season (set by `SEASON` below). The data requires extensive cleaning to standardize heights, positions, years, geographic information, and previous school names. Exploratory checks live in the **Data quality checks** section at the end.

## Setup and Configuration

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(postmastr)
library(usdata)
library(rvest)

# Configuration: change SEASON (e.g. "2027-28") and everything else follows
SEASON      <- "2026-27"
START_YEAR  <- as.integer(str_sub(SEASON, 1, 4))
SEASON_END  <- START_YEAR + 1L                                   # 2027
SEASON_TAG  <- str_replace(SEASON, "-", "_")                     # 2026_27
PREV_TAG    <- sprintf("%d_%02d", START_YEAR - 1L, START_YEAR %% 100L)  # 2025_26

INPUT_URL    <- paste0("https://raw.githubusercontent.com/dwillis/wbb-rosters/refs/heads/main/rosters_", SEASON, ".csv")
OUTPUT_FILE  <- paste0("wbb_rosters_", SEASON_TAG, ".csv")
PREV_FILE    <- paste0("wbb_rosters_", PREV_TAG, ".csv")
TEAMS_URL    <- "https://raw.githubusercontent.com/Sports-Roster-Data/womens-college-basketball/main/teams.csv"
POSITIONS_URL <- "https://raw.githubusercontent.com/dwillis/jour479x_fall_2022/main/data/wbb_positions.csv"
```

## Load Source Data

```{r load-data, message=FALSE}
# Load roster data and rename team_id column
wbb_rosters <- read_csv(INPUT_URL, show_col_types = FALSE) |>
  rename(ncaa_id = team_id) |>
  mutate(team = str_squish(team))

stopifnot(all(wbb_rosters$season == SEASON))
n_input <- nrow(wbb_rosters)

# Load teams metadata
teams <- read_csv(TEAMS_URL, show_col_types = FALSE) |>
  select(-url, -twitter, -stats_name) |>
  distinct()

# Join rosters with team metadata
wbb_rosters <- wbb_rosters |>
  left_join(teams, by = c("ncaa_id", "team"), relationship = "many-to-one")

stopifnot(nrow(wbb_rosters) == n_input)
```

## Height Cleaning

The roster data contains heights in various formats (6'0", 6-0, 6'0, etc.). We standardize all heights to use hyphens (6-0) and create a total inches column for analysis.

```{r clean-heights}
# Source heights come as 6'0", 6' 0'', 6-0, etc. Remove quotes and spaces, then split on - or '
wbb_rosters <- wbb_rosters |>
  mutate(
    height_clean = height |>
      str_remove_all("''|\"|\\s|(?i)height:") |>
      str_replace_all("'", "-")
  ) |>
  separate(height_clean, c("height_ft", "height_in"), sep = "-", extra = "merge", fill = "right") |>
  mutate(
    across(c(height_ft, height_in), \(x) suppressWarnings(as.numeric(x))),
    total_inches = height_ft * 12 + height_in,
    # Heights outside a plausible range are parse errors, not players
    total_inches = if_else(between(total_inches, 58, 90), total_inches, NA_real_)
  )
```

## Position Cleaning

Position data comes in various formats and combinations. We standardize positions and separate primary/secondary positions where applicable.

```{r clean-positions}
# Existing mapping from raw position strings to GUARD / FORWARD / CENTER / WING combinations
positions_cleaned <- read_csv(POSITIONS_URL, show_col_types = FALSE) |>
  select(-count) |>
  distinct(position, .keep_all = TRUE)

# Fallback for variants the mapping lacks (PG/G, SF/PF, "Position: Guard", ...):
# map each token to G/F/C/W, then rebuild in the same order the mapping uses.
token_map <- c(
  G = "G", PG = "G", SG = "G", CG = "G", GUARD = "G", GUARDS = "G", GUAD = "G", POINT = "G", SHOOTING = "G",
  F = "F", SF = "F", PF = "F", FW = "F", FWD = "F", FORWARD = "F", FOWARD = "F", P = "F", POWER = "F", SMALL = "F",
  C = "C", CENTER = "C", POST = "C", IF = "C",
  W = "W", WG = "W", WING = "W",
  `SMALL FORWARD` = "F"
)

fallback_position <- function(x) {
  map_chr(x, \(p) {
    if (is.na(p)) return(NA_character_)
    tokens <- unname(token_map[str_split_1(p, "\\s*[/-]\\s*")])
    if (anyNA(tokens)) return(NA_character_)
    if ("W" %in% tokens) return("WING")
    tokens <- intersect(c("G", "F", "C"), tokens)
    paste(recode(tokens, G = "GUARD", F = "FORWARD", C = "CENTER"), collapse = "/")
  })
}

wbb_rosters <- wbb_rosters |>
  mutate(position = str_squish(str_remove(position, regex("^Position:\\s*", ignore_case = TRUE)))) |>
  left_join(positions_cleaned, by = "position", relationship = "many-to-one") |>
  mutate(
    position_clean = coalesce(position_clean, fallback_position(str_to_upper(position))),
    position_full = position_clean
  ) |>
  # Split combined positions (e.g., "GUARD/FORWARD")
  separate(position_clean, c("primary_position", "secondary_position"),
           sep = "/", extra = "merge", fill = "right")

# Check position distribution
wbb_rosters |> count(primary_position, sort = TRUE)
```

## Year Classification Cleaning

We standardize year classifications to: Freshman, Sophomore, Junior, Senior, Graduate Student, Fifth Year, Sixth Year. We also track redshirt status separately.

```{r clean-years}
# Exact-match lookup for the known formats (one row per key)
years_cleaned <- read_csv("years_cleaned.csv", show_col_types = FALSE) |>
  rename(year_clean = `year-clean`) |>
  filter(!is.na(year_clean)) |>
  distinct(year, .keep_all = TRUE)

# Pattern fallback for everything the lookup misses (FRESHMAN, 2nd Year, Jr. (3rd), Red 5th, ...)
year_from_pattern <- function(x) {
  x <- x |> str_to_upper() |> str_squish() |> str_remove("^ACADEMIC |^R-|^RS-?|^RED ")
  case_when(
    str_detect(x, "^GR|GRAD")                       ~ "Graduate Student",
    str_detect(x, "SIXTH|SEVENTH|6TH")              ~ "Sixth Year",
    str_detect(x, "FIFTH|5TH")                      ~ "Fifth Year",
    str_detect(x, "^FR|FRESH|1ST|FIRST|^1$")        ~ "Freshman",
    str_detect(x, "^SO|SOPH|2ND|SECOND|^2$")        ~ "Sophomore",
    str_detect(x, "^JR|JUNIOR|3RD|THIRD|^3$")       ~ "Junior",
    str_detect(x, "^SR|SENIOR|4TH|FOURTH|^4$")      ~ "Senior",
    TRUE                                            ~ NA_character_
  )
}

# Class years in the source can also be graduation years (2030 / '30)
grad_year_class <- c("Freshman", "Sophomore", "Junior", "Senior")
grad_year_map <- tibble(
  year = c(as.character(SEASON_END + 3:0), paste0("'", (SEASON_END + 3:0) %% 100)),
  grad_year_clean = rep(grad_year_class, 2)
)

wbb_rosters <- wbb_rosters |>
  rename(any_of(c(year = "academic_year"))) |>
  left_join(years_cleaned, by = "year", relationship = "many-to-one") |>
  left_join(grad_year_map, by = "year", relationship = "many-to-one") |>
  mutate(
    year_clean = coalesce(year_clean, grad_year_clean, year_from_pattern(year)),
    redshirt = coalesce(
      redshirt,
      as.numeric(str_detect(str_to_upper(year), "REDSHIRT|^R-|^RS|^RED "))
    )
  ) |>
  select(-grad_year_clean)

# Check year distribution
wbb_rosters |> count(year_clean, sort = TRUE)
```

## High School and Previous School Cleaning

Some records list high schools in the previous_school column. We separate these and backfill missing high school data from previous roster years.

```{r clean-schools}
# Some records list high schools in the previous_school column
hs_pattern <- "HS|H\\.S\\.|High School"

previous_school_is_hs <- wbb_rosters |>
  filter(is.na(high_school), !is.na(previous_school), str_detect(previous_school, hs_pattern)) |>
  pull(previous_school)

wbb_rosters <- wbb_rosters |>
  mutate(
    hs_clean = if_else(is.na(high_school) & previous_school %in% previous_school_is_hs,
                       previous_school, high_school),
    # Clear previous_school when it is really the high school
    previous_school = if_else(previous_school == hs_clean, NA_character_, previous_school)
  )

# Backfill missing high schools from last season's file (same player, same team)
if (file.exists(PREV_FILE)) {
  previous_hs <- read_csv(PREV_FILE, show_col_types = FALSE) |>
    filter(!is.na(high_school)) |>
    distinct(name, team, ncaa_id, .keep_all = TRUE) |>
    select(name, team, ncaa_id, prev_high_school = high_school)

  wbb_rosters <- wbb_rosters |>
    left_join(previous_hs, by = c("name", "team", "ncaa_id"), relationship = "many-to-one") |>
    mutate(
      backfilled = is.na(hs_clean) & !is.na(prev_high_school),
      hs_clean = coalesce(hs_clean, prev_high_school)
    )

  cat("Backfilled", sum(wbb_rosters$backfilled), "high school records from", PREV_FILE, "\n")
  wbb_rosters <- select(wbb_rosters, -prev_high_school, -backfilled)
} else {
  cat("No previous season file", PREV_FILE, "- skipping high school backfill\n")
}
```

## Geographic Data Cleaning

We clean hometown/homestate data to standardize state abbreviations and identify international players.

```{r clean-geography}
# Standardize hometown format
wbb_rosters <- wbb_rosters %>%
  mutate(
    hometown_cleaned = hometown %>%
      str_to_upper() %>%
      str_replace_all("\\.", "") %>%
      str_replace("/.*", "") %>%
      str_replace("\\.$", "") %>%
      # Fix reversed format: "ILLINOIS, CHICAGO" -> "CHICAGO, ILLINOIS"
      str_replace("^ILLINOIS, CHICAGO$", "CHICAGO, ILLINOIS") %>%
      str_replace(",", ", ") %>%
      str_replace("-", " ") %>%
      str_squish()
  )

# Create extended state dictionary for postmastr
states_non_standard <- pm_append(
  type = "state",
  input = c("SD.", "MASS", "CALIF", "MICH", "NEB", "IND", "MINN", "ORE", "OHIO", 
            "FLA", "MISS", "TENN", "ARIZ", "KAN", "ALA", "OKLA", "WIS", "ILL", 
            "WASH", "ARK", "COLO", "NEV", "CONN", "WISC", "WVA", "DEL", "WYO",
            "CALI", "LOUIS", "VIRG", "MONT", "PENN", "TEX", "KANS", "NEBR", 
            "IDA", "COL"),
  output = c("SD", "MA", "CA", "MI", "NE", "IN", "MN", "OR", "OH", "FL", "MS", 
             "TN", "AZ", "KS", "AL", "OK", "WI", "IL", "WA", "AR", "CO", "NV", 
             "CT", "WI", "WV", "DE", "WY", "CA", "LA", "VA", "MT", "PA", "TX", 
             "KS", "NE", "ID", "CO"),
  locale = "us"
)

dict <- pm_dictionary(type = "state", case = c("title", "upper", "lower"), 
                     append = states_non_standard)

# Identify addresses and parse states using postmastr
wbb_rosters <- wbb_rosters %>%
  pm_identify(var = "hometown_cleaned")

parsed <- wbb_rosters %>%
  pm_prep(var = "hometown_cleaned", type = "street") %>%
  pm_state_parse(dict)

# Join parsed results
wbb_rosters <- wbb_rosters %>%
  left_join(parsed, by = "pm.uid") %>%
  separate(hometown, c("hometown", "homestate"), sep = ",", extra = "merge", fill = "right") %>%
  mutate(
    homestate = str_trim(homestate),
    homestate = str_replace(homestate, "\\.", ""),
    homestate = case_when(
      is.na(abbr2state(homestate)) ~ homestate,
      TRUE ~ abbr2state(homestate)
    )
  )

# Handle special state cases
wbb_rosters <- wbb_rosters %>%
  mutate(
    pm.state = case_when(
      homestate %in% c("Hawai'i", "O'ahu") ~ "HI",
      homestate == "Ken" ~ "KY",
      homestate %in% c("Caif", "Cal", "Calfiornia", "Califonia") ~ "CA",
      homestate %in% c("Illi", "IIl", "Illnois") ~ "IL",
      homestate == "Min" ~ "MN",
      homestate == "N C." ~ "NC",
      homestate == "NMex." ~ "NM",
      homestate == "Oreg" ~ "OR",
      homestate == "US.V.I." ~ "USVI",
      homestate == "Ver" ~ "VT",
      homestate == "Was" ~ "WA",
      homestate == "WS" ~ "WI",
      homestate == "NYC" ~ "NY",
      # Special cases
      is.na(pm.state) & is.na(homestate) & str_detect(hometown_cleaned, "^ARLINGTON$") ~ "TX",
      is.na(pm.state) & is.na(homestate) & str_detect(hometown_cleaned, "^LOS ANGELES$") ~ "CA",
      is.na(pm.state) & is.na(homestate) & str_detect(hometown_cleaned, "^GREENWOOD$") ~ "SC",
      is.na(pm.state) & is.na(homestate) & str_detect(hometown_cleaned, "^GWYNN PARK$") ~ "MD",
      is.na(pm.state) & is.na(homestate) & str_detect(hometown_cleaned, "^KING COVE$") ~ "AK",
      TRUE ~ pm.state
    )
  )
```

## Country Identification

The country is the last piece of the hometown unless a US state was parsed. Variant spellings, provinces and regions are standardized below.

```{r clean-countries}
# Extract potential country from last part of hometown
wbb_rosters <- wbb_rosters %>%
  mutate(
    country = if_else(!is.na(pm.state), "USA", NA_character_),
    # last comma-separated piece of the hometown (NA when hometown is missing/empty)
    temp = str_trim(str_extract(hometown_cleaned, "[^,]+$")),
    temp = if_else(temp %in% c("NA", ""), NA_character_, temp),
    # known nation names are kept as-is; anything else falls through to temp for standardization
    country = coalesce(country, temp)
  )

# Standardize country names
wbb_rosters <- wbb_rosters %>%
  mutate(
    country = case_when(
      # Canadian provinces
      country %in% c("BC", "QUEBEC", "BRITISH COLUMBIA", "ALBERTA", "ONTARIO", 
                    "NOVA SCOTIA", "ONT", "QUÉBEC", "MANITOBA", "QUE", "ON", 
                    "QC", "ALBERTA (AB)", "ONT CANADA", "BC CANADA", "BC,CANADA", 
                    "ON CANADA", "ALBERTA CANADA", "SASKATCHEWAN", "CN", "AB", 
                    "ALTA") ~ "CANADA",
      # Australian states
      country %in% c("VICTORIA", "SOUTH AUSTRALIA", "WESTERN AUSTRALIA", 
                    "TASMANIA", "AUSTRAILA", "AU", "AUS", 
                    "QUEENSLAND AUSTRALIA") ~ "AUSTRALIA",
      # UK variations
      country %in% c("GREAT BRITAIN", "ENGLAND", "ENG", "UK", "WALES", 
                    "NORTHERN IRELAND", "EAST MIDLANDS") ~ "UNITED KINGDOM",
      # Other standardizations
      country == "BARCELONA" ~ "SPAIN",
      country == "HAWAI'I" ~ "USA",
      country == "SWEEDEN" ~ "SWEDEN",
      country == 'HUDIKSVALL' ~ 'SWEDEN',
      country == 'KEFLAVIK ICELAND' ~ 'ICELAND',
      country == "CZECHIA" ~ "CZECH REPUBLIC",
      country == "DR" ~ "DOMINICAN REPUBLIC",
      country %in% c('THE NETHERLANDS', 'NETHLERLANDS', 'NOORD HOLLAND') ~ 'NETHERLANDS',
      country == 'MÉXICO' ~ 'MEXICO',
      country == 'NZ' ~ 'NEW ZEALAND',
      country == 'SÉNÉGAL' ~ 'SENEGAL',
      country %in% c('DR CONGO', 'DEMOCRATIC REPUBLIC OF CONGO', 'CONGO') ~ 'DEMOCRATIC REPUBLIC OF THE CONGO',
      country %in% c("TURKIYE", "TÜRKIYE") ~ "TURKEY",
      TRUE ~ country
    )
  ) %>%
  select(-temp)

```

## High School Name Standardization

Apply standardized high school names to reduce duplicates and improve consistency. See `HIGH_SCHOOL_STANDARDIZATION.md` for details.

```{r standardize-high-schools}
# Load high school mapping
hs_mapping_file <- "data/high_school_mapping.csv"

if (file.exists(hs_mapping_file)) {
  cat("Applying high school standardization...\n")

  hs_mapping <- read_csv(hs_mapping_file, show_col_types = FALSE) %>%
    select(high_school_original, high_school_standardized, confidence, source) %>%
    # A few schools appear twice; prefer the manually reviewed row
    arrange(confidence != "high_manual") %>%
    distinct(high_school_original, .keep_all = TRUE)

  cat(sprintf("  Loaded %d high school mappings\n", nrow(hs_mapping)))

  # Count unique schools before standardization
  unique_before <- n_distinct(wbb_rosters$hs_clean[!is.na(wbb_rosters$hs_clean)])

  # Apply mapping
  wbb_rosters <- wbb_rosters %>%
    left_join(
      hs_mapping,
      by = c("hs_clean" = "high_school_original"),
      relationship = "many-to-one"
    ) %>%
    mutate(
      # Use standardized name if available, otherwise keep hs_clean
      high_school_standardized = coalesce(high_school_standardized, hs_clean),
      # Set confidence level
      hs_confidence = case_when(
        !is.na(confidence) ~ confidence,
        country != "USA" ~ "international",
        is.na(hs_clean) ~ "missing",
        TRUE ~ "unstandardized"
      ),
      # Track if standardization was applied
      hs_was_standardized = hs_clean != high_school_standardized &
                           !is.na(high_school_standardized)
    ) %>%
    select(-confidence, -source)

  # Count unique schools after standardization
  unique_after <- n_distinct(wbb_rosters$high_school_standardized[!is.na(wbb_rosters$high_school_standardized)])

  # Report statistics
  total_hs <- sum(!is.na(wbb_rosters$hs_clean))
  standardized <- sum(wbb_rosters$hs_was_standardized, na.rm = TRUE)

  cat(sprintf("  Total records with high schools: %d\n", total_hs))
  cat(sprintf("  Standardized: %d (%.1f%%)\n",
              standardized,
              standardized / total_hs * 100))
  cat(sprintf("  Unique schools before: %d\n", unique_before))
  cat(sprintf("  Unique schools after: %d\n", unique_after))
  cat(sprintf("  Deduplication: %.1f%%\n",
              (unique_before - unique_after) / unique_before * 100))

  cat("\n  Confidence breakdown:\n")
  print(table(wbb_rosters$hs_confidence))

} else {
  cat("Note: High school mapping file not found:", hs_mapping_file, "\n")
  cat("      Run scripts/normalize_high_schools.py and scripts/build_hs_mapping.py to create it\n")
  cat("      Proceeding without standardization...\n")

  # Create placeholder columns
  wbb_rosters <- wbb_rosters %>%
    mutate(
      high_school_standardized = hs_clean,
      hs_confidence = "none",
      hs_was_standardized = FALSE
    )
}
```

## Previous School Standardization

We standardize previous school names to handle variations in how schools are referenced.

```{r standardize-previous-schools}
# Create standardization mapping for most common variations
previous_school_mapping <- tribble(
  ~previous_school, ~previous_school_clean,
  "Academic of Art University", "Academy of Art University",
  "Air Force Academy", "Air Force",
  "American Univ.", "American",
  "Arkansas Little Rock", "Arkansas-Little Rock",
  "Austin Peay State", "Austin Peay",
  "Cal Poly SLO", "Cal Poly",
  "CSU Bakersfield", "Cal State Bakersfield",
  "California Baptist University", "California Baptist",
  "ETSU", "East Tennessee State",
  "FDU", "Fairleigh Dickinson",
  "JMU", "James Madison",
  "LIU", "Long Island",
  "Louisiana State", "LSU",
  "Miami (FL)", "Miami",
  "Miami (OH)", "Miami Ohio",
  "N.C. State", "NC State",
  "Ole Miss", "Mississippi",
  "TCU", "Texas Christian",
  "UCF", "Central Florida",
  "UCSB", "UC Santa Barbara",
  "UIC", "Illinois-Chicago",
  "UMass", "Massachusetts",
  "USF", "South Florida",
  "UTEP", "Texas-El Paso",
  "VCU", "Virginia Commonwealth"
)

# Apply standardization
wbb_rosters <- wbb_rosters %>%
  left_join(previous_school_mapping, by = "previous_school", relationship = "many-to-one") %>%
  mutate(
    previous_school_clean = coalesce(previous_school_clean, previous_school)
  )

# Show transfer statistics
transfer_summary <- wbb_rosters %>%
  filter(!is.na(previous_school_clean)) %>%
  count(previous_school_clean, sort = TRUE) %>%
  head(10)

print(transfer_summary)
```

## Final Data Preparation

```{r final-cleanup}
wbb_rosters_final <- wbb_rosters |>
  select(-pm.id, -pm.uid, -pm.type, -pm.address) |>
  rename(
    height_clean = height,
    position_clean = position_full,
    hometown_clean = hometown_cleaned,
    state_clean = pm.state,
    country_clean = country
  ) |>
  distinct() |>
  # Same player listed twice on one roster (e.g. partial scrapes): keep the more complete row
  mutate(.filled = rowSums(!is.na(across(everything())))) |>
  slice_max(.filled, n = 1, by = c(ncaa_id, team, name, jersey), with_ties = FALSE) |>
  select(-.filled) |>
  relocate(team, ncaa_id, season, jersey, name, position, height_clean, year, hometown, homestate,
           high_school, previous_school, url, team_state, conference, division,
           height_ft, height_in, total_inches, primary_position, secondary_position,
           position_clean, year_clean, redshirt, hs_clean, hometown_clean, state_clean,
           country_clean, previous_school_clean) |>
  arrange(ncaa_id, jersey)

cat("Dropped", n_input - nrow(wbb_rosters_final), "duplicate player rows\n")
stopifnot(
  "rows were lost beyond duplicate players" = nrow(wbb_rosters_final) == n_distinct(wbb_rosters$ncaa_id, wbb_rosters$team, wbb_rosters$name, wbb_rosters$jersey),
  "wrong season" = all(wbb_rosters_final$season == SEASON)
)
```

## Manual Corrections

Edits made in the browser editor (`python editor/app.py --season 2026-27`) are logged in `corrections/corrections_<season>.csv`, one row per changed cell or deleted player. They are re-applied here on every knit. Players are matched on the as-scraped `ncaa_id`, `team`, `name` and `jersey`, so a corrected name doesn't break later corrections. The uncorrected output is saved as `corrections/baseline_<season>.csv`; the editor rebuilds the season file from it the same way.

```{r apply-corrections}
CORRECTIONS_FILE <- paste0("corrections/corrections_", SEASON_TAG, ".csv")
BASELINE_FILE    <- paste0("corrections/baseline_", SEASON_TAG, ".csv")

dir.create("corrections", showWarnings = FALSE)
write_csv(wbb_rosters_final, BASELINE_FILE, quote = "all", na = "")

if (file.exists(CORRECTIONS_FILE)) {
  corrections <- read_csv(CORRECTIONS_FILE, col_types = cols(.default = "c"), na = character())

  # Character copy of the baseline as written, so keys and old values compare as the editor sees them
  baseline_chr <- read_csv(BASELINE_FILE, col_types = cols(.default = "c"), na = "")
  key_of <- \(d) paste(d$ncaa_id, d$team, d$name, replace_na(d$jersey, ""), sep = "\t")
  corrections$row <- match(key_of(corrections), key_of(baseline_chr))
  corrections$row[corrections$action == "set" & !corrections$column %in% names(baseline_chr)] <- NA

  stale <- list()
  drop_rows <- integer()
  for (i in which(!is.na(corrections$row))) {
    cr <- corrections[i, ]
    if (cr$action == "delete_row") {
      drop_rows <- c(drop_rows, cr$row)
      next
    }
    current <- replace_na(baseline_chr[[cr$column]][cr$row], "")
    if (current != cr$old_value) stale[[length(stale) + 1]] <- mutate(cr, current = current)
    new_chr <- na_if(cr$new_value, "")
    baseline_chr[[cr$column]][cr$row] <- new_chr
    target <- wbb_rosters_final[[cr$column]]
    # An all-NA column is text as far as the editor is concerned
    if (all(is.na(target))) wbb_rosters_final[[cr$column]] <- target <- as.character(target)
    wbb_rosters_final[[cr$column]][cr$row] <- if (is.logical(target)) as.logical(new_chr)
                                              else if (is.numeric(target)) as.numeric(new_chr)
                                              else new_chr
  }
  if (length(drop_rows)) wbb_rosters_final <- wbb_rosters_final[-unique(drop_rows), ]

  unmatched <- filter(corrections, is.na(row))
  stale <- bind_rows(stale)
  cat(sprintf("Applied %d corrections from %s (%d unmatched, %d stale)\n",
              sum(!is.na(corrections$row)), CORRECTIONS_FILE, nrow(unmatched), nrow(stale)))
  # Player renamed or dropped upstream since the correction was made
  if (nrow(unmatched)) print(select(unmatched, team, name, jersey, action, column, new_value, note), n = Inf)
  # Scraped value changed since the correction was made; the correction was still applied
  if (nrow(stale)) print(select(stale, team, name, column, old_value, current, new_value, note), n = Inf)
} else {
  cat("No corrections file", CORRECTIONS_FILE, "\n")
}
```

## Write Output

```{r write-output}
write_csv(wbb_rosters_final, OUTPUT_FILE, quote = "all", na = "")

cat("Final dataset written to", OUTPUT_FILE)
cat("\nFinal statistics:")
cat("\n- Total players:", nrow(wbb_rosters_final))
cat("\n- Teams:", n_distinct(wbb_rosters_final$ncaa_id))
cat("\n- Transfer players:", sum(!is.na(wbb_rosters_final$previous_school_clean)))
cat("\n- International players:", sum(wbb_rosters_final$country_clean != "USA", na.rm = TRUE))
```

## Transfer Analysis by Conference

```{r transfer-analysis}
# Analyze transfer patterns for D1 teams
d1_transfer_analysis <- wbb_rosters_final %>%
  filter(division == "I") %>%
  group_by(team, conference) %>%
  summarise(
    total_players = n(),
    transfers = sum(!is.na(previous_school_clean)),
    .groups = "drop"
  ) %>%
  mutate(
    transfer_pct = transfers / total_players * 100
  ) %>%
  arrange(desc(transfer_pct))

# Conference-level transfer rates
conference_transfers <- d1_transfer_analysis %>%
  group_by(conference) %>%
  summarise(
    total_players = sum(total_players),
    total_transfers = sum(transfers),
    .groups = "drop"
  ) %>%
  mutate(
    transfer_pct = total_transfers / total_players * 100
  ) %>%
  arrange(desc(transfer_pct))

print("Top conferences by transfer percentage:")
print(head(conference_transfers, 20))
```


## Data quality checks

Run these after a render; anything that prints rows needs a look.

```{r check-coverage}
# Teams in teams.csv with no roster this season
teams |>
  anti_join(distinct(wbb_rosters_final, ncaa_id), by = "ncaa_id") |>
  select(ncaa_id, team, conference, division)

# Teams with players but no class year
wbb_rosters_final |>
  group_by(ncaa_id, team) |>
  summarize(n = n(), no_year = sum(is.na(year_clean)), .groups = "drop") |>
  filter(no_year == n)

# Division I roster sizes by conference
wbb_rosters_final |>
  filter(division == "I") |>
  count(conference, team, name = "players") |>
  group_by(conference) |>
  summarize(teams = n(), min = min(players), median = median(players), max = max(players))
```

```{r check-missing}
# NA counts for the cleaned columns
wbb_rosters_final |>
  summarize(across(c(total_inches, primary_position, year_clean, state_clean, country_clean,
                     hs_clean, previous_school_clean), \(x) sum(is.na(x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing")

# Source values that did not clean
wbb_rosters_final |> filter(!is.na(height_clean), is.na(total_inches)) |> count(height_clean, sort = TRUE)
wbb_rosters_final |> filter(!is.na(position), is.na(primary_position)) |> count(position, sort = TRUE)
wbb_rosters_final |> filter(!is.na(year), is.na(year_clean)) |> count(year, sort = TRUE)
```

```{r check-high-school-na}
# Teams with at least one player missing high_school, worst first
teams_missing_hs <- wbb_rosters_final |>
  filter(country_clean == "USA") |> 
  group_by(ncaa_id, team, conference, division) |>
  summarize(players = n(), missing_high_school = sum(is.na(high_school)), .groups = "drop") |>
  filter(missing_high_school > 0) |>
  mutate(pct_missing = round(missing_high_school / players * 100, 1)) |>
  arrange(desc(pct_missing), desc(missing_high_school))

teams_missing_hs
```

```{r check-duplicates}
# Same player listed twice on one roster
wbb_rosters_final |> count(ncaa_id, team, name, jersey) |> filter(n > 1)
```

```{r check-countries}
missing_countries <- wbb_rosters_final |> filter(is.na(country_clean))
```
