```{r}
library(tidyverse)
```

```{r}
wbb_rosters23 <- read_csv("wbb_rosters_2022_23.csv")
```

# Data Exploration Section

## Which countries outside the U.S. have supplied the most players?

Spain, Australia and Canada provide the most non-U.S. players by far.

```{r}
wbb_rosters23 %>% 
  filter(country_clean != 'USA') %>% 
  group_by(country_clean) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))
```




## Which teams/conferences/divisions have the most first-year players? Which have the most seniors & graduate students?

```{r}
teams_first_year <- wbb_rosters23 %>% filter(year_clean == "Freshman") %>% group_by(team, division) %>% summarise(count = n()) %>% arrange(desc(count))
#the top three schools are Mary Hardin Baylor, Simpson and Anderson (IN). The D1 school with the most freshmen is Northern Colorado

conference_first_year <- wbb_rosters23 %>% group_by(conference, division) %>% summarise(count = n()) %>% arrange(desc(count))
#PSAC, RMAC and NSIC have the most freshmen. All are D2 conferences. A10 is the fourth most, and the D1 conference with the most freshmen

division_first_year <- wbb_rosters23 %>% group_by(division) %>% summarise(count = n()) %>% arrange(desc(count))
#Division 1 had the most first year players, followed by D3 and then D2. 

teams_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(team, division) %>% summarise(count = n()) %>% arrange(desc(count))
#Austin Peay, Seton Hall and Gannon have the most seniors and grad students. The top two are D1 schools.

conference_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(conference, division) %>% summarise(count = n()) %>% arrange(desc(count))
#The ACC, SEC and and A10 have the most seniors and grad students. The top seven conferences by seniors and grad students are all D1 conferences. 

division_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(division) %>% summarise(count = n()) %>% arrange(desc(count))
#D1 has, by far, the most seniors and grad students. D1 has 1.6x times as many seniors and grad students as D3 (second most). 
```

#Data exploration: which teams/conferences/divisions have the most/least in-state players?
```{r}
team_in_state_players_na <- wbb_rosters23  %>%
  group_by(team, team_state, state_clean, division, conference) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  filter(is.na(state_clean))

team_recruiting <- wbb_rosters23 %>%
  group_by(team, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

conference_recruiting <- wbb_rosters23 %>%
  group_by(conference, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

division_recruiting <- wbb_rosters23 %>%
  group_by(division, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

wbb_rosters23 %>% group_by(team_state) %>% summarise(count = n_distinct(team)) %>% arrange(desc(count))

wbb_rosters23 %>% filter(is.na(team_state))
```

In a lot of cases, we got NA as a result for the state_clean. In the cases of the DI schools, this meant that someone was an international player. 