```{r}
library(tidyverse)
```

```{r}
wbb_rosters23 <- read_csv("wbb_rosters_2022_23.csv")
```

# Data Exploration Section

## Which countries outside the U.S. have supplied the most players?

Spain, Australia and Canada provide the most non-U.S. players by far.

```{r}
non_us <- wbb_rosters23 %>% 
  filter(country_clean != 'USA') %>% 
  group_by(country_clean) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))

non_us_division <- wbb_rosters23 %>% 
  filter(country_clean != 'USA') %>% 
  group_by(country_clean, division) %>% 
  summarize(players = n()) %>% 
  pivot_wider(names_from=division, values_from = players) %>% 
  replace(is.na(.), 0) %>% 
  mutate(total = I + II + III) %>% 
  mutate(pct_d1 = I/total*100)

non_us_team <- wbb_rosters23 %>% 
  filter(country_clean != 'USA') %>% 
  group_by(team) %>% 
  summarize(players = n()) %>% 
  arrange(desc(players))
```
More than 25 teams have at least seven non-U.S. players on their rosters, San Francisco has 11 and Portland and UT-Tyler have 10 each (Simon Fraser has 11 but it is located in Canada).

```{r}
wbb_rosters23 %>% 
  filter(country_clean != 'USA') %>% 
  group_by(team) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))
```



Of the 111 teams that have at least one Australian player, 37 of them have two or more on their roster. Portland WBB has _seven_ Aussies.

```{r}
wbb_rosters23 %>% 
  filter(country_clean == 'AUSTRALIA') %>% 
  group_by(team) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))
```




## Which teams/conferences/divisions have the most first-year players? Which have the most seniors & graduate students?

```{r}
teams_first_year <- wbb_rosters23 %>% filter(year_clean == "Freshman") %>% group_by(team, division) %>% summarise(count = n()) %>% arrange(desc(count))
#the top three schools are Mary Hardin Baylor, Simpson and Anderson (IN). The D1 school with the most freshmen is Northern Colorado

conference_first_year <- wbb_rosters23 %>% group_by(conference, division) %>% summarise(count = n()) %>% arrange(desc(count))
#PSAC, RMAC and NSIC have the most freshmen. All are D2 conferences. A10 is the fourth most, and the D1 conference with the most freshmen

division_first_year <- wbb_rosters23 %>% group_by(division) %>% summarise(count = n()) %>% arrange(desc(count))
#Division 1 had the most first year players, followed by D3 and then D2. 

teams_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(team, division) %>% summarise(count = n()) %>% arrange(desc(count))
#Austin Peay, Seton Hall and Gannon have the most seniors and grad students. The top two are D1 schools.

conference_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(conference, division) %>% summarise(count = n()) %>% arrange(desc(count))
#The ACC, SEC and and A10 have the most seniors and grad students. The top seven conferences by seniors and grad students are all D1 conferences. 

division_seniors_grad <- wbb_rosters23 %>% filter(year_clean == "Senior" | year_clean == "Graduate Student" | year_clean == "Sixth Year" | year_clean == "Fifth Year") %>% group_by(division) %>% summarise(count = n()) %>% arrange(desc(count))
#D1 has, by far, the most seniors and grad students. D1 has 1.6x times as many seniors and grad students as D3 (second most). 
```

#Data exploration: which teams/conferences/divisions have the most/least in-state players?
```{r}
team_in_state_players_na <- wbb_rosters23  %>%
  group_by(team, team_state, state_clean, division, conference) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  filter(is.na(state_clean))

team_recruiting <- wbb_rosters23 %>%
  group_by(team, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

conference_recruiting <- wbb_rosters23 %>%
  group_by(conference, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

division_recruiting <- wbb_rosters23 %>%
  group_by(division, team_state, state_clean) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

total_players <- wbb_rosters23 %>% 
  group_by(ncaa_id, team) %>% 
  summarize(all_players = n()) %>% 
  arrange(desc(all_players))

in_state_players <- wbb_rosters23 %>% 
  filter(team_state == state_clean) %>% 
  group_by(ncaa_id, team) %>% 
  summarize(in_state_players = n()) %>% 
  arrange(desc(in_state_players))

pct_in_state <- total_players %>% 
  left_join(in_state_players) %>% 
  mutate(in_state_players = ifelse(is.na(in_state_players),0, in_state_players))

wbb_rosters23 %>% group_by(team_state) %>% summarise(count = n_distinct(team)) %>% arrange(desc(count))

wbb_rosters23 %>% filter(is.na(team_state))
```

In a lot of cases, we got NA as a result for the state_clean. In the cases of the DI schools, this meant that someone was an international player. 

```{r}
no_hs <- wbb_rosters23 %>% filter(is.na(high_school), is.na(hs_clean), country_clean == 'USA')
```

