---
title: "Building Stable Cross-Season Player IDs"
output: html_document
---

This notebook builds `wbb_id`, a stable identifier that links the same
player across seasons and teams. The roster files' own `player_id` is just
the team-website roster number and changes every season (and is missing
entirely from 2025-26), so it cannot be used to track a player over time.
See `scripts/player_id_functions.R` for the matching logic; this notebook
runs the pipeline, writes the output files, and reports validation checks.

## How matching works

Each season is matched against a cumulative table of previously-seen
players, in three tiers:

1. **Tier 1 (returning players):** same normalized name on the same team,
   with a sane class-year progression (redshirts and the 2020-21 -> 2021-22
   COVID eligibility freeze allow a repeated class) and no height/state
   conflict.
2. **Tier 2 (transfers):** same normalized name on a *different* team,
   corroborated by `previous_school_clean` matching the prior team, or a
   matching hometown/high school.
3. **Tier 3 (fuzzy, review-only):** name-distance candidates blocked by team
   or by last-name + hometown/high-school. These are never auto-merged --
   they're written to `wbb_id_review_queue.csv` for a human decision.

Cross-team same-name pairs with **no** corroborating evidence that still pass the
same year and height screens are also queued now (`same_name_unconfirmed_transfer`),
as are corroborated pairs that fail a screen (`corroborated_blocked`, with the
evidence and the failing screen in their columns). Neither is ever auto-merged.

Anything left unmatched is a new person (freshman, transfer from outside the
dataset, or a genuine miss) and gets a freshly minted `wbb_id`.

Decisions recorded in `player_id_overrides.csv` (season/team/name pairs
marked `same` or `different`) are applied after matching and are honored on
every re-run, so manual adjudication persists.

## Setup

```{r setup, message=FALSE, warning=FALSE}
source("scripts/player_id_functions.R")
```

## Load and prepare all six seasons

```{r load}
all_rows <- load_all_seasons()
cat("Total rows loaded:", nrow(all_rows), "\n")
print(all_rows %>% count(season, season_order))

dedup <- resolve_within_season_dupes(all_rows)
all_rows_d <- dedup$rows
cat("\nWithin-season duplicate (same name + team) groups flagged as the same",
    "person (scraper double-listing):", sum(!is.na(all_rows_d$dup_of)), "\n")
cat("Within-season duplicate groups flagged as genuinely distinct players",
    "(sent to review queue):", nrow(dedup$review), "\n")
```

## Run the tiered matcher across all seasons

```{r match}
result <- build_player_ids(all_rows_d)

assignments <- propagate_dup_assignments(all_rows_d, result$assignments)
assignments <- apply_overrides(assignments, all_rows_d)

stopifnot(nrow(assignments) == nrow(all_rows_d))
cat("Assigned a wbb_id to all", nrow(assignments), "rows\n")
cat("Distinct players (wbb_id) across all seasons:", n_distinct(assignments$wbb_id), "\n")
```

## Validation

```{r validate}
val <- validate_ids(all_rows_d, assignments)

cat("Match-tier funnel by season:\n")
print(val$funnel)

if (length(val$issues) > 0) {
  warning("Validation issues found -- review before publishing:")
  print(val$issues)
} else {
  cat("\nNo validation issues: no wbb_id spans two teams in one season,",
      "and no wbb_id spans more than 7 seasons.\n")
}
```

### GUID crosswalk cross-check (2022-23)

`wbb_rosters23_crosswalk.csv` carries a persistent NCAA/Sidearm person GUID
for the 2022-23 season only. It can't link seasons by itself, but it is
strong independent ground truth for within-season distinctness: two rows we
assigned the same `wbb_id` should share a GUID (precision), and two rows
sharing a GUID should have gotten the same `wbb_id` (recall).

```{r guid-check}
guid_check <- check_against_guids(all_rows_d, assignments)
cat("Precision (same wbb_id -> same GUID):", round(guid_check$precision, 4), "\n")
cat("Recall (same GUID -> same wbb_id):", round(guid_check$recall, 4), "\n")
cat("Precision violations:", nrow(guid_check$precision_violations), "\n")
cat("Recall violations:", nrow(guid_check$recall_violations), "\n")
```

### Spot checks

```{r spot-checks}
joined <- all_rows_d %>% left_join(assignments, by = "row_id")

cat("Payton Hull (Abilene Christian) -- should be one wbb_id across three seasons:\n")
print(joined %>% filter(str_detect(name, "Payton Hull")) %>%
        select(season, team, name, season_player_id, wbb_id, match_tier) %>%
        arrange(season))

cat("\nKamryn Mraz (Abilene Christian) -- redshirt Sophomore -> redshirt Junior,",
    "should be one wbb_id:\n")
print(joined %>% filter(str_detect(name, "Kamryn Mraz")) %>%
        select(season, team, name, year_clean, redshirt, wbb_id, match_tier) %>%
        arrange(season))

cat("\nDaugherty sisters, Abilene Christian 2025-26 -- same team/hometown/high",
    "school but different jersey/height, must stay two distinct wbb_ids:\n")
print(joined %>% filter(str_detect(name, "Daugherty"), season == "2025-26") %>%
        select(season, team, name, jersey, total_inches_num, wbb_id, match_tier))
```

## Build output files

```{r build-players}
mode_value <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}

guid_map <- guid_check$joined %>%
  filter(!is.na(hhs_person_id_text)) %>%
  distinct(wbb_id, hhs_person_id_text) %>%
  group_by(wbb_id) %>%
  slice(1) %>%
  ungroup()

players <- joined %>%
  arrange(wbb_id, season_order) %>%
  group_by(wbb_id) %>%
  summarise(
    canonical_name = tail(name[order(season_order)], 1),
    first_season = head(season[order(season_order)], 1),
    last_season = tail(season[order(season_order)], 1),
    n_seasons = n_distinct(season),
    hometown_clean = mode_value(hometown_clean),
    state_clean = mode_value(state_clean),
    country_clean = mode_value(country_clean),
    hs_clean = mode_value(hs_clean),
    height_total_inches_mode = suppressWarnings(as.numeric(mode_value(total_inches_num))),
    position_mode = mode_value(position_clean),
    match_confidence = case_when(
      n_distinct(season) == 1 ~ "single_season",
      any(match_tier %in% c("tier2", "tier2_dup")) ~ "medium",
      TRUE ~ "high"
    ),
    .groups = "drop"
  ) %>%
  left_join(guid_map, by = "wbb_id") %>%
  rename(hhs_person_id = hhs_person_id_text)

write_csv(players, "players.csv", quote = "all", na = "")
cat("Wrote players.csv:", nrow(players), "players\n")
print(table(players$match_confidence))
```

```{r build-player-seasons}
player_seasons <- joined %>%
  arrange(wbb_id, season_order) %>%
  select(wbb_id, season, ncaa_id, team, name, season_player_id, url,
         year_clean, redshirt, match_tier)

write_csv(player_seasons, "wbb_player_seasons.csv", quote = "all", na = "")
cat("Wrote wbb_player_seasons.csv:", nrow(player_seasons), "rows\n")
```

```{r build-combined}
wbb_rosters_combined <- joined %>%
  select(-name_norm, -last_name_norm, -url_norm, -year_rank, -redshirt_num,
         -total_inches_num, -dup_of, -is_dup_representative, -row_id,
         -season_order, -match_tier) %>%
  arrange(season, ncaa_id, jersey)

stopifnot(nrow(wbb_rosters_combined) == nrow(all_rows_d))
stopifnot(sum(is.na(wbb_rosters_combined$wbb_id)) == 0)

write_csv(wbb_rosters_combined, "wbb_rosters_combined.csv", quote = "all", na = "")
cat("Wrote wbb_rosters_combined.csv:", nrow(wbb_rosters_combined), "rows\n")
```

```{r build-review-queue}
review_queue_final <- bind_rows(result$review_queue, dedup$review) %>%
  mutate(decision = NA_character_) %>%
  arrange(desc(score))

write_csv(review_queue_final, "wbb_id_review_queue.csv", quote = "all", na = "")
cat("Wrote wbb_id_review_queue.csv:", nrow(review_queue_final), "candidate rows",
    "for manual review\n")
```

```{r overrides-template}
if (!file.exists("player_id_overrides.csv")) {
  tibble(
    season_a = character(), team_a = character(), name_a = character(),
    season_b = character(), team_b = character(), name_b = character(),
    decision = character(), note = character()
  ) %>% write_csv("player_id_overrides.csv", quote = "all")
  cat("Created empty player_id_overrides.csv template\n")
} else {
  cat("player_id_overrides.csv already exists -- left untouched\n")
}
```

## Maintenance loop for a new season

1. Add the new season's file to `SEASON_FILES` in
   `scripts/player_id_functions.R` and re-knit this notebook.
2. Open `wbb_id_review_queue.csv`, fill in the blank `decision` column
   (`same` or `different`) for rows you can resolve, and copy those
   decisions into `player_id_overrides.csv` using the natural-key columns
   (`season_a/team_a/name_a` vs `season_b/team_b/name_b`).
3. Re-knit. Overrides are applied on every run, so adjudicated decisions
   persist across seasons.
